---
title: "dplyr Introduction"
author: "CB"
date: "Tuesday, May 9, 2017"
output:
  html_document:
    keep_md: yes
---
```{r include = FALSE}
knitr::opts_chunk$set(cache = TRUE, autodep = TRUE)
```

## Chapter 5

### 5.1 Introduction

dplyr can be very useful for transforming data into a form that is easier to work with. 

#### 5.1.1 Prerequisites

Load dplyr and ggplot2, which are both included in tidyverse. Also load nycflights13, which includes data on 2013 flights from New York City.

```{r}
library(nycflights13)
library(tidyverse)
```

There are some conflicts between dplyr and base R functions, but the base functions can still be used by specifying them from the stats package. For example, you can specify the base `filter()` by using `stats::filter()`. 

#### 5.1.2 nycflights13

This dataset is set up as a tibble, which will display differently than a standard data frame.
```{r}
flights
```

The variable type is listed under each column header and specifies the format of that column. 

* int = integers
* dbl = real numbers
* chr = characters
* dttm = date-times
* lgl = logical
* fctr = factors
* date = dates

#### 5.1.3 dplyr Basics

Main dplyr functions

* `filter()`: select observations by values
* `arrange()`: reorder rows
* `select()`: select variables by names
* `mutate()`: make new variables from existing variables
* `summarise()`: create a summary of many values
* `group_by()`: limits a function to work on a specific group instead of the entire dataset

Provided a data frame and instructions for modifying the data frame, these functions will create a new data frame as specified.

### 5.2 `filter()`

Use `filter()` to select all flights on January 1st.
```{r}
filter(flights, month == 1, day == 1)
```

Using parentheses around a variable assignment will both save the result to a variable and show the result.
```{r}
(dec25 <- filter(flights, month == 12, day == 25))
```

#### 5.2.1 Comparisons

R includes a language for comparing two things.

* `>`: greater than
* `>=`: greater than or equal to
* `<`: less than
* `<=`: less than or equal to 
* `!=`: not equal 
* `==`: equal

When working with floating point numbers, `near()` can be more accurate than `==`. 
```{r}
1/49 * 49 == 1
near(1/49 * 49, 1)
```

#### 5.2.2 Logical Operators

Arguments can be combined and only data that satisfies all requirements will be included in the output.

* `&`: and
* `|`: or
* `!`: not

Select all flights departing in November or December.
```{r}
filter(flights, month == 11 | month == 12)
```

`x %in% y` can alternatively be used to select values of x in y.
```{r}
filter(flights, month %in% c(11, 12))
```

There are often multiple ways to write these comparisons. For example, here are two ways to find flights that weren't delayed by more than two hours.
```{r}
filter(flights, !(arr_delay > 120 | dep_delay > 120))
filter(flights, arr_delay <= 120, dep_delay <= 120)
```

#### 5.2.3 Missing Values

The result of almost any function involving NA will be NA. If missing values should be included, specifically ask for them. `is.na()` can be used to check if a value is missing.

```{r}
df <- tibble(x = c(1, NA, 3))
filter(df, x >1)
filter(df, is.na(x) | x > 1)
```

#### 5.2.4 Exercises

1. 
  1. Flights with an arrival delay of two or more hours
```{r}
filter(flights, arr_delay >= 120)
```

  2. Flights to Houston (IAH or HOU)
```{r}
filter(flights, dest %in% c("IAH", "HOU"))
```

  3. Flights operated by United, American, or Delta
```{r}
filter(flights, carrier %in% c("UA", "AA", "DL"))
```

  4. Flights that departed in July, August, or September
```{r}
filter(flights, month %in% c(7, 8, 9))
```

  5. Flights that arrived over two hours late but didn't leave late
```{r}
filter(flights, dep_delay <= 0 & arr_delay > 120)
```

  6. Flights delayed by at least an hour that made up over 30 minutes in flight
```{r}
filter(flights, dep_delay > 60 & arr_delay < 30)
```

  7. Flights that departed between midnight and 6 am (including these times)
```{r}
filter(flights, hour %in% c(12, 1, 2, 3, 4, 5, 6))
```

2. `between()` specifies a range of values for a variable and can be used to simplify challenge 4 and 7.
```{r}
filter(flights, between(month, 7, 9))
filter(flights, hour == 12 | between(hour, 1, 6))
```

3. There are 8,255 flights with a missing `dep_time`. These rows may represent cancelled flights. **What other variables are missing?**
```{r}
nrow(filter(flights, is.na(dep_time)))
```

4. `NA ^ 0` is not missing because anything to the 0 power = 1. `NA | TRUE` is not missing because it is converted into a TRUE or FALSE value. `FALSE & NA` is not missing because it is also converted into a TRUE or FALSE value. If NA is converted into some other value through an equation, the result will not be missing.

### 5.3 `arrange()`

`arrange()` reorders rows by specified columns, where additional columns are used to break ties in previous columns. `desc()` arranges a column in descending order with missing values at the end.

```{r}
df <- tibble(x = c(5, 2, NA)) 
arrange(df, x)
arrange(df, desc(x))
```

#### 5.3.1 Exercises

1. Sort missing values to the start with `arrange()`
```{r}
arrange(df, desc(is.na(x)))
```

2. Find flights that were most delayed and flights that left earliest
```{r}
# most delayed
arrange(flights, desc(dep_delay))

# left earliest
arrange(flights, arr_delay)
```

3. Find the fastest flights
```{r}
arrange(flights, air_time)
```

4. Find flights that travelled longest and shortest
```{r}
# longest
arrange(flights, desc(distance))

# shortest
arrange(flights, distance)
```

### 5.4 `select()`

`select()` is useful for limiting variables (columns) to those you're interested in when working with a large dataset.

Additional functions can be used within `select()`

* `starts_with("abc")`: selects names beginning with "abc"
* `ends_with("xyz")`: selects names ending with "xyz"
* `contains("ijk")`: selects names containing "ijk"
* `matches("(.)\\1")`: selects variables matching a regular expression (here, containing repeated characters)
* `num_range("x", 1:3)`: selects names matching x1, x2, and x3

`rename()` can be used to rename specific variables.
`select()` with `everything()` is useful for keeping all variables but moving specific variables to the beginning of the data frame.

#### 5.4.1 Exercises

1. Select dep_time, dep_dealy, arr_time, and arr_delay in as many ways as possible.
```{r}
select(flights, dep_time, dep_delay, arr_time, arr_delay)
select(flights, starts_with("arr"), starts_with("dep"))
```

2. `select()` ignores repeated variables and only includes them once in the output.

3. `one_of()` will select all variables specified in a vector. It would be helpful with the `vars` vector to select all columns invluded in that vector. 
```{r}
vars <- c("year", "month", "day", "dep_delay", "arr_delay")
select(flights, one_of(vars))
```

4. The following code was surprising because it included all variables with "time" regardless of upper and lower case. By default, the select helpers ignore case. The default can be changed by specifying `ignore.case = FALSE`.
```{r}
select(flights, contains("TIME"))
```

### 5.5 `mutate()`

`mutate()` creates new variables using existing variables and adds them to the end of the dataset.

Make a smaller dataset to easier see the effects of `mutate()`.
```{r}
flights_sml <- select(flights,
                      year:day,
                      ends_with("delay"),
                      distance,
                      air_time)

mutate(flights_sml,
       gain = arr_delay - dep_delay,
       speed = distance / air_time * 60)
```

`transmute()` will only keep new variables. 

#### 5.5.1

Functions that can be used with `mutate()` must be able to operate on vectors and return vector outputs.

Standard arithmetic functions as well as modular arithmetic functions are useful with `mutate()`. 
```{r}
transmute(flights,
          dep_time,
          hour = dep_time %/% 100,
          minute = dep_time %% 100)
```

Logarithmic functions can also be used with `mutate()`. `lead()` refers to leading values and `lag()` refers to lagging values. 

`min_rank()` will rank values within a column.

#### 5.5.2 Exercises

1. Convert dep_time and sched_dep_time to minutes since midnight. Need to parse hours and minutes?
```{r}

```

2. I would expect air_time to match arr_time - dep_time, but this is not the case.
```{r}
transmute(flights, 
          air_time,
          arr_dep = arr_time - dep_time)
```

3. I would expect dep_delay to equal dep_time - sched_dep_time.
```{r}
transmute(flights,
          dep_time,
          sched_dep_time,
          dep_delay,
          delay = dep_time - sched_dep_time)
```

4. 
```{r}
mutate(flights_sml, 
       min_rank(desc(dep_delay)))
```

5. 1:3 is looped since it is longer than 1:10.
```{r}
1:3 + 1:10
```

6. R provides cosine, sine, tangent, arc-cosine, arc-since, arc-tangent, and two-argument arc-tangent as trigonometric functions.