---
title: "Strings and Regular Expressions cont."
author: "CB"
date: "Tuesday, July 18, 2017"
output:
  html_document:
    keep_md: yes
---
```{r include = FALSE}
knitr::opts_chunk$set(cache = TRUE, autodep = TRUE)
library(tidyverse)
library(stringr)
```

## Chapter 14 Strings

### 14.4 Tools

#### 14.4.1 Detect matches

`str_detect()` can find matches within character vectors.
```{r}
x <- c("apple", "banana", "pear")
str_detect(x, "e")
```

Logical vectors are useful for counting matches.
```{r}
sum(str_detect(words, "^t"))
mean(str_detect(words, "[aeiou]$"))
```

Regular expressions can be broken up into smaller pieces to make them easier to work with and understand rather than using one long and complex regular expression.

`str_detect()` within a subsetting function works the same as `str_subset()`.
```{r}
words[str_detect(words, "x$")]
str_subset(words, "x$")
```

`str_detect()` can also be used in conjunction with `filter()`.
```{r}
df <- tibble(
  word = words,
  i = seq_along(word)
)
df %>% 
  filter(str_detect(words, "x$"))
```

`str_count()` is like `str_detect()` but counts the number of matches within the string.
```{r}
x <- c("apple", "banana", "pear")
str_count(x, "a")
```

`str_count()` can be used with `mutate()`. 
```{r}
df %>% 
  mutate(
    vowels = str_count(word, "[aeiou]"),
    consonants = str_count(word, "[^aeiou]")
  )
```

String matches don't overlap.
```{r}
str_count("abababa", "aba")
str_view_all("abababa", "aba")
```

Many stringr functions are paired so that one function works with one match and the other function works with all matches.

#### 14.4.2 Exercises

##### 1

1. all words that start or end with `x`
```{r}
# single regular expression
(single <- str_subset(words, "^x|x$"))

# multiple str_detect() calls
x_start <- str_detect(words, "^x")
x_end <- str_detect(words, "x$")
(multiple <- words[x_start | x_end])

identical(single, multiple)
```

2. all words that start with a vowel and end with a consonant
```{r}
# single regular expression
single <- str_subset(words, "^[aeiou].*[^aeiou]$")
head(single)

# multiple str_detect() calls
vowel_start <- str_detect(words, "^[aeiou]")
consonant_end <- str_detect(words, "[^aeiou]$")
multiple <- words[vowel_start & consonant_end]
head(multiple)

identical(single, multiple)
```

3. words that contain at least one of each vowel
```{r}
# single regular expression (from stackoverflow)
vowels <- "^(?=.*a)(?=.*e)(?=.*i)(?=.*o)(?=.*u).*$"
(single <-  str_subset(words, vowels))

# multiple str_detect() calls
a <- str_detect(words, "a")
e <- str_detect(words, "e")
i <- str_detect(words, "i")
o <- str_detect(words, "o")
u <- str_detect(words, "u")
(multiple <- words[a & e & i & o & u])

identical(single, multiple)

# did they work?
test <- c("easiloud", "aeiou", "around")
str_subset(test, vowels)

a_test <- str_detect(test, "a")
e_test <- str_detect(test, "e")
i_test <- str_detect(test, "i")
o_test <- str_detect(test, "o")
u_test <- str_detect(test, "u")
test[a_test & e_test & i_test & o_test & u_test]
```

##### 2

What word has the highest number of vowels?
```{r}
df <- tibble(
  word = words
)

df %>% 
  mutate(
    vowels = str_count(word, "[aeiou]")
  ) %>% 
  arrange(desc(vowels)) %>% 
  head()

# all words with 5 (most) vowels
df %>% 
  mutate(
    vowels = str_count(word, "[aeiou]")
  ) %>% 
  arrange(desc(vowels)) %>% 
  filter(vowels == 5)
```

What word has the highest proportion of vowels? 
```{r}
df %>% 
  mutate(
    vowels = str_count(word, "[aeiou]"),
    vowel_proportion = vowels/str_length(word)
  ) %>% 
  arrange(desc(vowel_proportion)) %>% 
  head()
```

#### 14.4.3 Extract matches

`str_extract()` will extract matched text, but only the first match.
```{r}
colors <- c("red", "orange", "yellow", "green", "blue", "purple")
(color_match <- str_c(colors, collapse = "|"))
has_color <- str_subset(sentences, color_match)
matches <- str_extract(has_color, color_match)
head(matches)
```

Use `str_extract_all()` to extract all matches.
```{r}
more <- sentences[str_count(sentences, color_match) > 1]
str_view_all(more, color_match)
str_extract_all(more, color_match)
```

`simplify = TRUE` will return a matrix instead of a list.
```{r}
x <- c("a", "a b", "a b c")
str_extract_all(x, "[a-z]", simplify = TRUE)
```

##### 14.4.3.1 Exercises

###### 1 

Only match colors and not colors within other words.
```{r}
color_match <- "\\b(red|orange|yellow|green|blue|purple)\\b"

more <- sentences[str_count(sentences, color_match) > 1]
str_view_all(more, color_match)
```

###### 2

1. Extract the first word from each sentence
```{r}
str_extract(sentences, "^\\w+") %>% head()
```

2. Extract all words ending in `ing`
```{r}
ing_words <- "\\w+ing\\b"
has_ing <- str_subset(sentences, ing_words)
str_extract_all(has_ing, ing_words) %>% head()
```

3. Extract all plurals
```{r}
plurals <- "\\w{2,}[b-hj-np-rtv-z]s\\b"
has_plurals <- str_subset(sentences, plurals)
str_extract_all(has_plurals, plurals) %>% head()
```

#### 14.4.4 Grouped matches

Extract nouns from sentences.
```{r}
noun <- "(a|the) ([^ ]+)"

has_noun <- sentences %>% 
  str_subset(noun) %>% 
  head(10)

has_noun %>% 
  str_extract(noun)
```

`str_match()` splits matches into groups.
```{r}
has_noun %>% 
  str_match(noun)
```

`tidyr::extract()` is easier to use with tibbles.
```{r}
tibble(sentence = sentences) %>% 
  tidyr::extract(
    sentence, c("article", "noun"), "(a|the) ([^ ]+)",
    remove = FALSE
  )
```

##### 14.4.4.1 Exercises

1. Pull out numbers and the words that come after them.
```{r}
number_word <- "\\b(one|two|three|four|five|six|seven|eight|nine|ten) (\\w+)"
has_nw <- sentences %>% 
  str_subset(number_word)
has_nw %>% 
  str_match_all(number_word) %>% 
  head()
```

2. Find all contractions and separate the parts around the apostrophe.
```{r}
contraction <- "([A-Za-z]+)'([a-z]+)"
has_con <- sentences %>% 
  str_subset(contraction)
has_con %>% 
  str_match_all(contraction) %>% 
  head()
```

#### 14.4.5 Replacing matches

`str_replace()` will replace matches with something new.
```{r}
x <- c("apple", "pear", "banana")
str_replace(x, "[aeiou]", "-")
str_replace_all(x, "[aeiou]", "-")

x <- c("1 house", "2 cars", "3 people")
str_replace_all(x, c("1" = "one", "2" = "two", "3" = "three"))
```

Backreferences can also be used for replacements.
```{r}
sentences %>% 
  str_replace("([^ ]+) ([^ ]+) ([^ ]+)", "\\1 \\3 \\2") %>% 
  head(3)
```

##### 14.4.5.1 Exercises

1. Replace forward slashes with backslashes
```{r}
x <- c("rat/mouse", "cat/dog", "bird/bat")
str_replace(x, "/", "\\")
str_replace(x, "/", "\\\\")
# can't get just one backslash as a replacement
```

2. Use `replace_all()` for `str_to_lower()`
```{r}
# not sure how to translate to the correct lowercase character
#sentences %>% 
 # str_replace_all("([A-Z])", "") %>% 
  #head()
```

3.  Switch first and last letter in `words`
```{r}
head(words)
str_replace(words, "^(\\w)(\\w*)(\\w)$", "\\3\\2\\1") %>% 
  head()
```
WOrds with the same first and last letter would still be words.

#### 14.4.6 Splitting

`str_split()` will split strings into pices. 
```{r}
sentences %>% 
  head(3) %>% 
  str_split(" ")

x <- "This is a sentence. This is another sentence."
str_view_all(x, boundary("word"))
```

##### 14.4.6.1 Exercises

1. Split a string into individual components
```{r}
x <- "apples, pears, and bananas"
str_split(x, boundary("word"))
```

2. Splitting by `boundary("word")` will remove punctuation when separating words, while splitting by `" "` includes the punctuation with the words they are next to.

3. Splitting with an empty string splits on each character. 
```{r}
str_split(x, "")
```

#### 14.4.7 Find matches

`str_locate()` gives the positions of each match, which is useful for dealing with situations where other functions aren't working well.

### 14.5 Other types of pattern

`regexp()` parameters can control pattern matching.

`fixed()` matches exactly what is specified on a byte level. This may be faster than `regexp()` for simple matches.
`coll()` matches strings with collation rules and uses `locale` to compare characters. This is slower than `regexp()` and `fixed()`. 

#### 14.5.1 Exercises

1. To find all strings containing `\` with `regex()`, you would need to escape the backslash and make sure it is being recognized as a literal backslash. This would be easier to find with `fixed()` because it could recognize `\` without the need for escaping the special character.

2. Five most common words in `sentences`
```{r}
words <- str_split(sentences, boundary("word"), simplify = TRUE)
# how to match all words that are the same and then count?
```

