---
title: "Strings and Regular Expressions"
author: "CB"
date: "Tuesday, July 12, 2017"
output:
  html_document:
    keep_md: yes
---
```{r include = FALSE}
knitr::opts_chunk$set(cache = TRUE, autodep = TRUE)
library(tidyverse)
```

## Chapter 14 Strings

### 14.1 Introduction

Regular expressions describe patterns in strings.

```{r}
library(stringr)
```

### 14.2 String basics

```{r}
string1 <- "This is a string"
string2 <- 'If I want to include a "quote" inside a string, I use single quotes'

double_quote <- "\"" # or '"'
single_quote <- '\'' # or "'"
backslash <- "\\"
```

The string is different from its printed representation because of escape characters.
```{r}
x <- c("\"", "\\")
x
writeLines(x)
```

Special characters

* `"\n"`: new line
* `"\t"`: tab

`?"'"` will show all special characters. There are also strings to write non-English characters.
```{r}
x <- "\u00b5"
x
```

#### 14.2.1 String length

```{r}
str_length(c("a", "R for data science", NA))
```

#### 14.2.2 Combining strings

Strings can be combined and the separator can be specified.
```{r}
str_c("x", "y")
str_c("x", "y", sep = ", ")
```

`str_replace_na()` can be used to print NA values as "NA". 
```{r}
x <- c("abc", NA)
str_c("|-", x, "-|")
str_c("|-", str_replace_na(x), "-|")
```

Objects of length 0 are dropped.
```{r}
name <- "Cassandra"
time_of_day <- "evening"
birthday <- FALSE

str_c(
  "Good ", time_of_day, " ", name,
  if (birthday) " and HAPPY BIRTHDAY",
  "."
)
```

Vectors of strings can be collapsed into single strings.
```{r}
str_c(c("x", "y", "z"), collapse = ", ")
```

#### 14.2.3 Subsetting strings

Parts of a string can be extracted. 
```{r}
x <- c("Apple", "Banana", "Pear")
str_sub(x, 1, 3)
str_sub(x, -3, -1)
str_sub("a", 1, 5)
```

`str_sub()` has an assignment form to modify strings.
```{r}
str_sub(x, 1, 1) <- str_to_lower(str_sub(x, 1, 1))
x
```

#### 14.2.4 Locales

Capitalization and sorting are affected by the locale. `str_sort()` and `str_order()` take additional locale arguments for robust analyses. 

```{r}
str_sort(x, locale = "en")
```

#### 14.2.5 Exercises

1. `paste()` uses space as a default separator while `paste0()` does not have a default separator. `paste()` and `paste0()` are equivalent to `str_c()`. `paste()` treats NA like a character whereas `str_c()` gives NA as the output if NA is in the input.
```{r}
paste("x", "y", NA)
paste0("x", "y", NA)
str_c("x", "y", NA)
```

2. `sep` is inserted between the arguments of `str_c()` while `collapse` is inserted between elements of a vector within `str_c()`. 
```{r}
# separate arguments 
str_c("x", "y", sep = ":")
str_c("x", "y", collapse = ":")
# vector
str_c(letters, sep = ":")
str_c(letters, collapse = ":")
# both
str_c(1:26, letters, sep = " ", collapse = ", ")
```

3. It looks like when an even string is divided by 2, R takes the first of the two middle characters. Maybe you could write an ifelse command to take both middle characters if the string has an even number of characters?
```{r}
x <- c("apple", "banana", "pear")
str_length(x)
str_sub(x, str_length(x)/2, str_length(x)/2)
```

4. `str_wrap()` wraps strings into a target width per line. This could be used to modify string visualization for printing onto paper or to fit better on a computer screen.

5. `str_trim()` will remove extra spaces from the start and/or end of strings. `str_pad()` is the opposite of `str_trim()` and will add whitespace to the sides of a string.

6. 

## regexone 

### Tutorial

space: â£
tab: \t
new line: \n
carriage return: \r
any whitespace: \s
any non-whitespace: \S
any digits: \d
any non-digits: \D
letters and digits: \w
non-alphanumeric characters: \W
any number of: *
any number of, but at least 1: +
optional: ?
different possible characters: |
any character: .
boundary between word and non-word characters: \b
reference captured groups: \0 for entire match, \1 for first group

1: abc
1.5: 123
2: ...\.
3: [cmf]an
4: [^b]og
5: [A-Z][a-z][a-z]
6: waz{2,5}up
7: a+b*c+
8: [1-9]+ files? found\?
9: [1-9]\.\s+abc
10: [A-Za-z]+:[^un]successful$
11: ^(\w+)\.pdf$
12: ^([A-Z][a-z]{2}\s(\d{4}))$
13: ^(\d{3}\d?)x(\d{3}\d?)$
14: ^I love (cats|dogs)$
15: ^.+\.$

### Practice problems

