---
title: "dplyr Continued + Script Workflow"
author: "CB"
date: "Tuesday, May 16, 2017"
output:
  html_document:
    keep_md: yes
---
```{r include = FALSE}
knitr::opts_chunk$set(cache = TRUE, autodep = TRUE)
library(tidyverse)
library(nycflights13)
```

## Chapter 5

### 5.6 `summarise()`

`summarise()` is used to summarize data and is most useful with `group_by()`. Together, these commands will create a grouped summary where the data is summarized by specified groups. 

```{r}
summarise(flights, delay = mean(dep_delay, na.rm = TRUE))

by_day <- group_by(flights, year, month, day)
summarise(by_day, delay = mean(dep_delay, na.rm = TRUE))
```

#### 5.6.1 Combining multiple operations with the pipe

Investigate the relationship between distance and average delay by location with and without using the pipe. 

**Without the pipe:**
```{r}
by_dest <- group_by(flights, dest)
delay <- summarise(by_dest,
                   count = n(),
                   dist = mean(distance, na.rm = TRUE),
                   delay = mean(arr_delay, na.rm = TRUE))
delay <- filter(delay, count > 20, dest != "HNL")

ggplot(delay, aes(x = dist, y = delay)) + 
  geom_point(aes(size = count), alpha = 1/3) + 
  geom_smooth(se = FALSE)
```

In this process, the flights are grouped by destination; distance, average delay, and number of flights are calculated with `summarise()`; and noisy points and Honolulu airport are removed with `filter()`. 

**With the pipe**
```{r}
delays <- flights %>% 
  group_by(dest) %>% 
  summarise(
    count = n(),
    dist = mean(distance, na.rm = TRUE),
    delay = mean(arr_delay, na.rm = TRUE)
  ) %>% 
  filter(count > 20, dest != "HNL")
```

Using the pipe makes code more readable and also simplifies the flow of code because you don't have to name intermediate variables between each function. Instead, the pipe feeds the result of the previous function into the next function. 

However, the pipe cannot be used within ggplot2 as it was developed after ggplot2. 

**`Ctrl` + `Shift` + `M` is a shortcute for the pipe. `%>%`**

#### 5.6.2 Missing values

Aggregation functions include missing values if there are any missing values in the input, so it may be helpful to remove these from the output by specifying `na.rm = TRUE`. 

```{r}
flights %>% 
  group_by(year, month, day) %>% 
  summarise(mean = mean(dep_delay))

flights %>% 
  group_by(year, month, day) %>% 
  summarise(mean = mean(dep_delay, na.rm = TRUE))
```  

Alternatively, you can remove the problematic observations first and create a new dataset without NA values. 

```{r}
not_cancelled <- flights %>% 
  filter(!is.na(dep_delay), !is.na(arr_delay))

not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(mean = mean(dep_delay))
```

#### 5.6.3 Counts

Including a count or count of non-missing values with aggregation functions is useful to check the amount of data included in the result. 

```{r}
delays <- not_cancelled %>% 
  group_by(tailnum) %>% 
  summarise(
    delay = mean(arr_delay)
  )

ggplot(delays, aes(x = delay)) + 
  geom_freqpoly(binwidth = 10)

delays <- not_cancelled %>% 
  group_by(tailnum) %>% 
  summarise(
    delay = mean(arr_delay, na.rm = TRUE),
    n = n()
  )

ggplot(delays, aes(x = n, y = delay)) + 
  geom_point(alpha = 1/10)
```

Here you can see that variation in flight delay is high when there are few flights used for the calculation. This can be fixed by filtering out groups with a small number of flights.

```{r}
delays %>% 
  filter(n > 25) %>% 
  ggplot(aes(x = n, y = delay)) + 
  geom_point(alpha = 1/10)
```

**`Ctrl` + `Shift` + `P` will re-run the last chunk of code.**

#### 5.6.4 Useful summary functions

R includes many summary functions

* Measures of location: `mean()` and `median()`

  Aggregation functions can also be combind with logical subsetting. 
```{r}
not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(avg_delay1 = mean(arr_delay),
            avg_delay2 = mean(arr_delay[arr_delay > 0])
  )
```

* Measures of spread: `sd()`, `IQR()`, and `mad()`

  `IQR()` gives the interquartile range and `mad()` gives the median absolute deviation. These may be more useful than `sd()` in situations with outliers. 
```{r}
not_cancelled %>% 
  group_by(dest) %>% 
  summarise(distance_sd = sd(distance)) %>% 
  arrange(desc(distance_sd))
```

* Measures of rank: `min()`, `quantile()`, and `max()`

  `quantile()` is similar to `median()` but can be specified. For instance, `quantile(x, 0.25)` will give a value of x greater than 25% of the values and less than 75% of the values. 
```{r}
not_cancelled %>%
  group_by(year, month, day) %>% 
  summarise(
    first = min(dep_time),
    last = max(dep_time)
  )
```

* Measures of position: `first()`, `nth()`, and `last()`

  `nth(x, 2)` will provide the second value of x. 
```{r}
not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(
    first_dep = first(dep_time),
    last_dep = last(dep_time)
  )
```

  Filtering by rank produces all variables and observations in the output while creating a summary of position only produces the involved variables and specified observations in the output.
```{r}
not_cancelled %>% 
  group_by(year, month, day) %>% 
  mutate(r = min_rank(desc(dep_time))) %>% 
  filter(r %in% range(r))
```

* Counts: `n()`, `sum(!is.na())`, and `n_distinct()`

  `n()` gives the size of the current group, `sum(!is.na())` gives a count of non-missing values, and `n_distinct()` gives a count of unique values. 
```{r}
not_cancelled %>% 
  group_by(dest) %>% 
  summarise(carriers = n_distinct(carrier)) %>% 
  arrange(desc(carriers))
```
  `count()` is also a specific function for counts and can include a weight variable to count with. 
```{r}
not_cancelled %>% 
  count(dest)

not_cancelled %>% 
  count(tailnum, wt = distance)
```

* Counts and proportions of logical values: `sum(x > 10)` and `mean(y == 0)`
  `TRUE` is converted to 1 and `FALSE` is converted to 0. So `sum()` will give a count of a logical value and `mean()` will give a proportion of a logical value.
```{r}
not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(n_early = sum(dep_time < 500))

not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(hour_perc = mean(arr_delay > 60))
```

#### 5.6.5 Grouping by multiple variables

`summarise()` will apply to the last level of a multiple variable grouping, which allows for progressive calculations by group. 

```{r}
daily <- group_by(flights, year, month, day)
(per_day <- summarise(daily, flights = n()))
(per_month <- summarise(per_day, flights = sum(flights)))
(per_year <- summarise(per_month, flights = sum(flights)))
```

However, be careful with group analysis because some progressive calculations by group will not equal the overall calculation without group. Sums and counts can be progressively calculated accurately while the median cannot. 

#### 5.6.6 Ungrouping

`ungroup()` will remove grouping from the data. 
```{r}
daily %>% 
  summarise(flights = n())

daily %>% 
  ungroup() %>% 
  summarise(flights = n())
```

#### 5.6.7 Exercises

1. Arrival delay is probably more important than departure delay because a plane could leave late and then make up time in the air and arrive on time. 
```{r}
# Separately looking at mean arrival and mean departure delays
not_cancelled %>% 
  group_by(dest, carrier) %>% 
  summarise(mean_arr_delay = mean(arr_delay),
            mean_dep_delay = mean(dep_delay))

# Looking at delays in the air
not_cancelled %>% 
  mutate(air_delay = arr_delay - dep_delay) %>% 
  group_by(dest, carrier) %>% 
  summarise(mean_air_delay = mean(air_delay))

# Separately looking at positive and negative delays for arrival and departure
not_cancelled %>% 
  group_by(dest, carrier) %>% 
  summarise(mean_pos_arr_delay = mean(arr_delay > 0),
            mean_neg_arr_delay = mean(arr_delay < 0),
            mean_pos_dep_delay = mean(dep_delay > 0),
            mean_neg_dep_delay = mean(dep_delay < 0))
```

2. Want same output as below:
```{r}
# original
not_cancelled %>% 
  count(dest)

# new
not_cancelled %>% 
  group_by(dest) %>% 
  summarise(n = n())

# original
not_cancelled %>% 
  count(tailnum, wt = distance)

# new
not_cancelled %>% 
  group_by(tailnum) %>% 
  summarise(n = sum(distance))
```

3. If a flight is canceled, it will neither depart nor arrive. However, a flight could depart and then not arrive if it crashes, but this scenario would not accurately fall under a canceled flight category. So the `dep_delay` column is most important for canceled flights.

4. In general, the proportion of canceled flights does relate to the average delay where a high proportion of canceled flights correlates with a high average delay. However, this relationship breaks down at the higher extremes of canceled flights.
```{r}
# Canceled flights per day
(canceled <- flights %>% 
  group_by(year, month, day) %>% 
  summarise(canceled = sum(is.na(dep_time)),
            prop_canceled = sum(is.na(dep_time)) / n(),
            avg_delay = mean(dep_delay, na.rm = TRUE))) 

# Proportion of canceled flights and average delay
ggplot(canceled, aes(x = prop_canceled, y = avg_delay)) + 
  geom_point() + 
  geom_smooth(se = FALSE)

# Zoom in to bottom left corner
canceled %>% 
  filter(prop_canceled < 0.3, avg_delay < 60) %>% 
  ggplot(aes(x = prop_canceled, y = avg_delay)) + 
  geom_point() + 
  geom_smooth(se = FALSE)
```

5. Carrier F9 has the worst average departure delays.
```{r}
not_cancelled %>% 
  group_by(carrier) %>% 
  summarise(mean_dep_delay = mean(dep_delay)) %>% 
  arrange(desc(mean_dep_delay))
```

6. `sort` will arrange the output of `count()` in descending order when `sort = TRUE` is specified. The default is `sort = FALSE`. This could be used to determine ranking of a specified count. For example, `sort` could be used to display the top ten days with the most canceled flights.

### 5.7 Grouped mutates and filters

In addition to being used with `summarise()`, `group_by()` can be useful with `mutate()` and `filter()`. 

* Worst members of each group
```{r}
flights_sml <- select(flights,
                      year:day,
                      ends_with("delay"),
                      distance,
                      air_time)

flights_sml %>% 
  group_by(year, month, day) %>% 
  filter(rank(desc(arr_delay)) < 10)
```

* Groups bigger than a threshold
```{r}
(popular_dests <- flights %>% 
  group_by(dest) %>% 
  filter(n() > 365))
```

* Calculate by group
```{r}
popular_dests %>% 
  filter(arr_delay > 0) %>% 
  mutate(prop_delay = arr_delay / sum(arr_delay)) %>% 
  select(year:day, dest, arr_delay, prop_delay)
```

Be careful with grouped filters because they are hard to check. Instead, you could use a grouped mutate and then an ungrouped filter.
```{r}
flights_sml %>% 
  group_by(year, month, day) %>% 
  filter(rank(desc(arr_delay)) < 10)

flights_sml %>% 
  group_by(year, month, day) %>% 
  mutate(rank = rank(desc(arr_delay))) %>% 
  ungroup() %>% 
  filter(rank < 10)
```

Window functions work best with grouped mutates and filters. These functions will provide n outputs for n inputs instead of one output for n inputs like summary functions.

#### 5.7.1 Exercises

1. 